---
title: "Student Performance Analysis"
author: "Rasheed Johnson"
format: 
  html:
    theme: flatly
    toc: true
    toc-location: left
execute:
  warning: false
  message: false
---

## Purpose

Frank Johnson High School wants to understand how variables such as "Study Hours", "Smartphone usage", "Subject", and "Learning Method" affect the exam scores of their students. This would help the school determine what interventions or programs they should develop to continue supporting their students' development.

## Variables Examined

-   **Study Hours/Day:** The amount of study hours per student
-   **Attendance %:** Attendance percentage per student
-   **Exam Score:** General exam scores per student (Doesn't specify a subject)
-   **Favorite Subject:** Students' favorite subjects\
-   **Smartphone Usage:** The amount of hours students use their smartphones
-   **Learning Method:** The method students attend school (e.g., Hybrid, On-site, Online)

## Analysis Goals

1.  Is there a relationship between smartphone usage and student exam scores? (Correlation Test)
2.  Is there a significant mean difference in exam scores between STEM & Non-STEM Majors? (T-Test)
3.  Can attendance predict student exam scores? (Linear Regression)
4.  Do students who learn On-site perform better on exams than students who are online? (T-Test)

## Hypotheses

-   **H1:** There will be a strong negative relationship between student smartphone usage and student exam scores.
-   **H2:** Students in STEM Majors will have a significantly higher mean in exam scores compared to Non-STEM Majors.
-   **H3:** Student attendance will be a strong positive predictor of student exam scores.
-   **H4:** Students who attend school in person will have a significantly higher mean in exam scores compared to students who learn online.

# 1. INSTALL & LOAD PACKAGES

```{r}
#| echo: true
#| warning: false
#| message: false
#| output: false
my_packages <- c("tidyverse", "stats", "effectsize", "ggtext")
lapply(my_packages, require, character.only = TRUE)
```

# 2. LOAD, VIEW DATA, & CHECK FOR NA'S

```{r}
students_df <- read_csv("education_dataset.csv")
glimpse(students_df)
summary(students_df)
colSums(is.na(students_df))
```

# 3. DATA CLEANING FOR STATISTICAL TEST

## 3.1 Set up for correlation and linear regression test

```{r}
relevant_cols <- c("Study Hours/Day", 
                   "Attendance %",
                   "Exam Score",
                   "Favorite Subject",
                   "Smartphone Usage (hrs/day)",
                   "Learning Method")

analyzed_students_df <- students_df %>% 
  select(all_of(relevant_cols)) 

glimpse(analyzed_students_df)
```

## 3.2 Set up for Independent Samples T-Test

```{r}
analyzed_students_df <- analyzed_students_df %>% 
  mutate(
    Majors = case_when(
      `Favorite Subject` == "Math" ~ 1,
      `Favorite Subject` == "Computer" ~ 1,
      `Favorite Subject` == "Science" ~ 1,
      `Favorite Subject` == "English" ~ 0,
      `Favorite Subject` == "History" ~ 0
    ) %>% factor(
      levels = c(0,1),
      labels = c("Non-STEM", "STEM")
    )
  )

Lmethod_t_test <- analyzed_students_df %>% 
  filter(`Learning Method` %in% c("Online", "Offline")) %>% 
  mutate(
    Learning_Group = case_when(
      `Learning Method` == "Online" ~ 1,
      `Learning Method` == "Offline" ~ 0
    ) %>% 
      factor(
        levels = c(1,0),
        labels = c("Online", "Onsite")
      )
  )

# Viewing if sample size is balanced for the t-test
analyzed_students_df %>% count(Majors) 
Lmethod_t_test %>% count(Learning_Group)
```

# 4. EXPLORATORY DATA ANALYSIS (VISUALIZATIONS)

## 4.1 Visualize the distribution for each single variable being analyzed

```{r}
ggplot(analyzed_students_df, aes(x = `Study Hours/Day`)) + 
  geom_histogram(color = "black", fill = "pink")
  
ggplot(analyzed_students_df, aes(x = `Attendance %`)) +
  geom_histogram(bins = 40, color = "black", fill = "blue") 

ggplot(analyzed_students_df, aes(x = `Exam Score`)) +
  geom_histogram(color = "black", fill = "lightblue") 

ggplot(analyzed_students_df, aes(x = `Smartphone Usage (hrs/day)`)) + 
  geom_histogram(color = "black", fill = "darkred")
```

## 4.2 Visualize scatter plots for H1 & H3

```{r}
ggplot(analyzed_students_df, aes(x = `Smartphone Usage (hrs/day)`,
                                 y = `Exam Score`)) +
  geom_point(alpha = 0.5, color = "black", size = 2) +
  geom_smooth(method = "lm", color = "red", linewidth = 1, se = TRUE) +
  labs(x = "Students Smart Phone Usage (Hrs)",
       y = "Students Exam Score")

ggplot(analyzed_students_df, aes(x = `Attendance %`,
                                 y = `Exam Score`)) + 
  geom_point(color = "black", size = 3) + 
  geom_smooth(method = "lm", se = TRUE, linewidth = 1) + 
  labs(x = "Students Attendance",
       y = "Students Exam Score")
```

## 4.3 Visualize T-Test for H2 & H4

```{r}
ggplot(analyzed_students_df, aes(x = Majors,
                                 y = `Exam Score`)) + 
  geom_boxplot(aes(fill = Majors))

ggplot(Lmethod_t_test, aes(x = Learning_Group,
                           y = `Exam Score`)) +
  geom_boxplot()
```

# 5. STATISTICAL TEST FOR ANALYSIS

## 5.1 Conduct Correlation Statistical Test

```{r}
cor.test(analyzed_students_df$`Smartphone Usage (hrs/day)`, 
         analyzed_students_df$`Exam Score`,
         alternative = "less",
         method = "pearson",
         conf.level = 0.95)
```

::: {.callout-note title="Key Finding"}
While there was a statistically significant finding between student smartphone usage & student exam score, the effect size was very low (*r* = -0.07). Meaning that there is no practical important relationship between student smartphone usage & student exam score, the significant result can be mostly due to the large sample size rather than real world impact.
:::

## 5.2 Conduct Linear Regression Statistical Test

```{r}
regression_model <- lm(`Exam Score` ~ `Attendance %`,
                       data = analyzed_students_df)
summary(regression_model)
```

::: {.callout-note title="Key Finding"}
There was not a statistically significant finding between student attendance & student exam scores. The predictive strength and direction was found to be very weak (*b* = -0.08). Student attendance could not be used as a variable to predict student exam performance (*r2* = 0.00).
:::

## 5.3 Conduct Independent Sample T-Test Statistical Tests

```{r}
t.test(`Exam Score` ~ Majors, 
       var.equal = FALSE, 
       conf.level = 0.95,
       data = analyzed_students_df)

descrpitive_stats <- analyzed_students_df %>% 
  group_by(Majors) %>% 
  summarise(
    mean = mean(`Exam Score`),
    sd = sd(`Exam Score`),
    n = n()
  )

mean_difference <- descrpitive_stats[2, 2] - descrpitive_stats[1, 2]
cohens_d <- cohens_d(`Exam Score` ~ Majors, 
                     data = analyzed_students_df)

descrpitive_stats
mean_difference
cohens_d
```

::: {.callout-note title="Key Finding"}
There was a statistically significant mean difference in exam scores between STEM & Non-STEM majors (p \< 0.05). Although STEM majors scored an average of 4.31 points higher in exam scores compared to Non-STEM majors, the effect size was small (*d* = 0.23). This means that the difference in exam scores between both groups overlap a lot.
:::

```{r}
t.test(`Exam Score` ~ Learning_Group,
       var.equal = FALSE,
       data = Lmethod_t_test,
       conf.level = 0.95)

lm_descrpitive_stats <- Lmethod_t_test %>% 
  group_by(Learning_Group) %>% 
  summarise(
    mean = mean(`Exam Score`),
    sd = sd(`Exam Score`),
    N = n()
  )

lm_mean_difference <- lm_descrpitive_stats[2, 2] - lm_descrpitive_stats[1, 2]
lm_cohen_d <- cohens_d(`Exam Score` ~ Learning_Group,
         data = Lmethod_t_test)

lm_descrpitive_stats
lm_mean_difference
lm_cohen_d
```

::: {.callout-note title="Key Finding"}
There was not a statistically significant mean difference in exam score between Online & On-Site students (p \> 0.09). Although the Onsite students scored an average of 3.63 points higher in exam score compared to Online students, the effect size was small. Implying that the difference in exam scores between both groups overlap.
:::

# 6. FINAL VISUALIZATIONS

## 6.1 Correlation visualization of student smart phone usage and exam score

```{r}
ggplot(analyzed_students_df, aes(x = `Smartphone Usage (hrs/day)`,
                                 y = `Exam Score`)
       ) +
  geom_point(color = "#6200EE", size = 2, alpha = 0.5) + 
  geom_smooth(method = "lm", color = "#03DAC5", se = TRUE, linewidth = 1.5) +
  labs(
    title = "No Practical Relationship between Student Smartphone Usage and Exam Scores",
    subtitle = "Relationship found statistically significant (*p* < 0.05) but, the effect size is low (*r* = -0.07). Significance susceptible to large sample size instead of real world effects.",
    x = "Smartphone Usage (Hours)",
    y = "Exam Scores") + 
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", margin = margin(b = 10)),
    plot.subtitle = element_textbox_simple(margin = margin(b = 15)),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    axis.title = element_text(size = 11),
    axis.line = element_line(linetype = "dashed"),
    axis.text = element_text(size = 10)) 
```

## 6.2 Linear regression visualization of student attendance and exam scores

```{r}
ggplot(analyzed_students_df, aes(x = `Attendance %`, y = `Exam Score`)
       ) +
  geom_jitter(width = 0.5, height = 0.5, alpha = 0.5, color = "black") +
  geom_smooth(method = "lm", color = "red", se = TRUE, linewidth = 1.5) +
  labs(x = "Attendance (Percentage)",
       title = "Student Attendance Unable to Predict Student Exam Scores ",
       subtitle = "Relationship was not statistically significant and had weak predictive strength and direction (*b* = -0.08). Student attendance not a strong variable to predict attendance (*r2* = 0.00).") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", margin = margin(b = 5)),
        plot.subtitle = element_textbox_simple(margin = margin(b = 15)),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 10),
        axis.line = element_line()) 
```

## 6.3 Independent samples t-test visualization (STEM & Non-STEM Majors)

```{r}
ggplot(analyzed_students_df, aes(x = Majors, y = `Exam Score`)
       ) +
  geom_boxplot( aes( fill = Majors)) +
  labs(title = "No Strong Difference in Exam Scores between Majors",
       subtitle = "Although the difference was statistically significant (*p* < 0.05),
       the effect size was small (*d* = 0.23).") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", margin = margin(b = 10)),
        plot.subtitle = element_textbox_simple(margin = margin(b = 15)),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(linetype = "dashed"),
        axis.text = element_text(size = 10))
```

## 6.4 Independent samples t-test visualization (Online & Onsite Students)

```{r}
ggplot(Lmethod_t_test, aes(x = Learning_Group, y = `Exam Score`)
       ) +
  geom_boxplot(aes(fill = Learning_Group)) +
  scale_fill_brewer(palette = "BuPu") + 
  labs(title = " No Strong Difference in Exam Scores between Learning Groups",
       subtitle = "The difference was not statistically significant (*p* > 0.05) and the effect size was small (*d* = -0.19). Implying that the difference in exam scores between both groups over lap.") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", margin = margin(r = 10, b = 5)),
        plot.subtitle = element_textbox_simple(margin = margin(t = 10, b = 12)),
        panel.grid = element_blank(),
        axis.line = element_line(linetype = "dashed"),
        axis.text = element_text(size = 10)
  )
```

# 7. Conclusions

::: {.callout-note title="Final Remarks"}
Considering there were not strong effects found in the chosen variables on exam scores when assessed through statistical tests. Frank Johnson high school should experiment with new variables to assess how the new variables may affect their student academic performance leading to the potential development of a initiative or investment in a program that will help the school support the improvement of their students.
:::
